Engineering an infinite-range SU($n$) spin model with ultracold fermions in an optical lattice

### 2021-04-22 sun_phases/spin_model_paper/main.tex
updated paper in light of feedback from AMR and RLS

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\Braket} % shorthand for braket notation
\renewcommand{\v}{\bm} % bold vectors
\renewcommand{\c}{\cdot} % inner product
\renewcommand{\d}{\partial} % partial derivative
\renewcommand{\dd}{\text{d}} % for infinitesimals
\renewcommand{\i}{\mathrm{i}\mkern1mu} % imaginary unit

@@

% more shorthands
\newcommand{\up}{\uparrow}
\newcommand{\dn}{\downarrow}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}

\newcommand{\xx}{\x\x}
\newcommand{\xxi}{\x\x_\i}
\newcommand{\X}{\text{X}}
\newcommand{\XX}{\X\X}
\newcommand{\XXI}{\X\X_\i}
\newcommand{\spin}{\text{spin}}
\newcommand{\eff}{\text{eff}}
\newcommand{\crit}{\text{crit}}
\newcommand{\MF}{\text{MF}}
\newcommand{\su}{\mathfrak{su}}
\renewcommand{\ss}{\bar{\v s}\c\bar{\v s}}
\DeclareMathOperator{\diag}{diag}

@@

% for text markup
\newcommand{\red}[1]{{\color{red} #1}}

@@

\begin{abstract}
We study multilevel fermions in an optical lattice, with dynamics described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime, this system can be mapped onto a multilevel Heisenberg spin model with all-to-all SU($n$)-symmetric couplings.
The use of Raman pulses to address internal spin states modifies the atomic dispersion relation and induces spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the non-equilibrium mean-field dynamical phases of the resulting model as a function of $n$ and different initial configurations that are accessible with the Raman pulses.
In agreement with previous studies with $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
When $n>2$, we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU(2) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU(2) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, the consideration of disordered SU($n$) spin models has opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless, bentsen2019integrable}.

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, beverland2016realizing, cazalilla2014ultracold, taie2012su, hofrichter2016direct, cappellini2014direct, scazza2014observation, zhang2014spectroscopic, goban2018emergence, perlin2019effective} with exquisite degrees of microscopic control.
In the simplest example, this symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ in the case of ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, cazalilla2014ultracold}.
As a result, experiments can directly probe the role SU($n$) interactions in controllable settings.
Recent progress includes the experimental observations of bosonization and faster thermalization in multi-component Fermi gas with $SU(n)$ symmetry \cite{song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, two-orbital SU($n$) magnetism \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic}, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.
In the spirit of quantum simulation, further investigations in controlled settings will play an important role in understanding SU($n$) symmetries' consequences for fundamental questions in physics, as well as their practical use in technological applications.
For example, SU(2)-symmetric spin interactions can be harnessed to develop quantum sensors that surpass classical limits on measurement precision \cite{he2019engineering, perlin2020spin}.
The prospect of similarly exploiting larger SU($n$) symmetries to achieve a technological advantage is still an unexplored avenue of research with untapped potential.

In this work, we consider an experimentally relevant and theoretically tractable regime of the SU($n$) Hubbard model, highlighting differences and similarities with the more familiar case of SU(2).
Working at ultracold temperatures and unit spatial filling (one atom per lattice site), we begin by mapping the SU($n$) Hubbard model onto a multilevel spin model with all-to-all SU($n$)-symmetric interactions in Section \ref{sec:spin_model}.
We consider the effect of spin-orbit coupling (SOC) in Section \ref{sec:SOC}, finding in particular that the weak-SOC limit generally gives rise to a (synthetic) inhomogeneous magnetic field, extending previously known results to $n>2$ \cite{wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
We take a look at control fields to address nuclear spins (and generate SOC) in Section \ref{sec:controls}, finding a simple three-laser driving scheme that allows for the preparation of interesting states with nontrivial intra-spin correlations when $n>2$.
Putting all of these ingredients together, we examine mean-field dynamical behaviors of the SU($n$) spin model in Section \ref{sec:mean_field}, finding that
\begin{enumerate*}
\item long-time-averaged observables obey simple scaling relations with the spin dimension $n$, exhibiting (for spin-polarized initial states) a magnetized and a de-magnetized dynamical phase, as previously seen for the case of $n=2$ \cite{smale2019observation, lewis-swan2021cavityqed}, and
\item when $n>2$ long-time dynamics can be highly sensitive to the intra-spin coherences of the initial state.
\end{enumerate*}
We conclude and discuss future outlooks in Section \ref{sec:conclusions}.

@@

Here we review the realization of a collective SU($n$) spin model in a system of ultracold alkaline-earth(-like) atoms on an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by single-body kinetic and two-body interaction Hamiltonians of the form
\begin{align}
  H_{\t{kin}}
  &= -\f{J}{2} \sum_{\bk{j,j'},\mu} c_{j\mu}^\dag c_{j'\mu} + \t{h.c.}, \\
  H_{\t{int}}
  &= \f{U_0}{2} \sum_{j,\mu,\nu}
  c_{j\mu}^\dag c_{j\mu} c_{j\nu}^\dag c_{j\nu},
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude; and $U_0$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which may become relevant in a sufficiently shallow lattice, namely when $J\gtrsim E_{\t{R}}$, where $E_{\t{R}}$ is the atomic lattice recoil energy.
For simplicity, we now assume a periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$, $c_{j\mu}=\frac1{\sqrt{L}}\sum_k e^{-\i q\c j} c_{q\mu}$, finding that
\begin{align}
  H_{\t{kin}}
  &= -J\sum_{q,\mu} \cos\p{q} c_{q\mu}^\dag c_{q\mu}, \\
  H_{\t{int}}
  &= \f{U}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  c_{k\mu}^\dag c_{\ell\mu} c_{p\nu}^\dag c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $U\equiv U_0\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum), and we work in units with lattice spacing $a=1$.

If the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, then the mode-changing collisions in $H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work)\footnote{Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $c_{\pi-p,\mu}^\dag c_{\pi-q,\mu} c_{p\nu}^\dag c_{q\nu}$, which conserve both momentum and energy.
We defer a careful treatment of these terms to future work, noting only that they vanish on the manifold of permutationally symmetric spin states with one atom per lattice site.}.
The operator content of terms with $k=\ell$ and $p=q$ takes the form $c_{\ell\mu}^\dag c_{\ell\mu} c_{q\nu}^\dag c_{q\nu}$, so these terms contribute an overall energy $\frac12NU$ that can be neglected in the absence of coherent atom number fluctuations.
Defining the spin operators $s_{\mu\nu q}\equiv c_{q\mu}^\dag c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians become
\begin{align}
  H_{\t{kin}} &= -J\sum_{q,\mu} \cos\p{q} s_{\mu\mu q},
  \label{H_kin_start} \\
  H_{\t{int}} &= -\f{U}{2N} \sum_{p,q,\mu,\nu} s_{\mu\nu p} s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case, we can neglect the fermionic origins of $H_{\t{int}}$ and instead treat our system as a bona fide spin model.
Note that the ``kinetic'' terms of this spin model ($H_{\t{kin}}$) are proportional to the identity operator, and can be neglected for the time being.
The fermionic origin of these terms will return to play an important role when we consider spin-orbit coupling in Section \ref{sec:SOC}.

To further simplify the interaction Hamiltonian $H_{\t{int}}$ and write it in a form suggestively similar to SU(2) spin alignment, we now construct the operator-valued spin matrix
\begin{align}
  \v s_q \equiv \sum_{\mu,\nu} s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v A,\v B$, we define the inner product
\begin{align}
  \v A \c \v B \equiv \Tr_{\t{aux}}\p{\v A^\dag\v B}
  = \sum_{\mu,\nu} A_{\mu\nu}^\dag B_{\mu\nu},
  \label{eq:dot_product}
\end{align}
where the trace is performed over the auxiliary matrix degrees of freedom.
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  H_{\t{int}} = -\f{U}{2N} \sum_{p,q} \v s_p\c\v s_q
  = -\f{U}{2N}\v S\c\v S,
  &&
  \v S \equiv \sum_q \v s_q.
  \label{eq:H_int}
\end{align}
Here $\v S$ is a collective spin matrix, analogous to the collective spin vector $\vec S=\p{S_\x,S_\y,S_\z}$ in the case of SU(2) \cite{he2019engineering}, with $\v S\c\v S \simeq 2 \vec S\c\vec S = 2 \sum_{\alpha\in\set{\x,\y,\z}}S_\alpha^2$ when $n=2$ (here $\simeq$ denotes equality after modding out identity terms).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites can tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U_0$.
(b,c) When the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, the frozen-mode approximation enables writing the interaction Hamiltonian as a spin model consisting of exchange terms $\v s_p\c\v s_q$, which swap the states two spins pinned to momentum modes $p,q$.
(d) Interactions thereby open an energy gap $U=U_0\times N/L$ between the manifold of permutationally symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v s_p\c\v s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v s_p\c\v s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v s_p\c\v s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns (distinct) definite energies to anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\v s_p\c\v s_q$, the interaction Hamiltonian $H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $U$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

In the case of SU(2), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

External fields or additional interactions that respect permutational symmetry can induce nontrivial dynamics within the PS manifold.
Even if additional Hamiltonian terms break permutational symmetry, as long as the coupling to states outside of the PS manifold is small compared to the interaction energy $U$, these terms can be restricted to the PS manifold perturbatively (see Appendix \ref{sec:pert_theory}) \cite{bravyi2011schrieffer}.
Simulating dynamics within the PS manifold requires calculating matrix elements $\bk{\ell|\O|m}$ of spin operators $\O$ with respect to PS states $\ket\ell,\ket{m}$; we discuss this calculation in Appendix \ref{sec:PS_ops}.

>>>>

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\braket} % shorthand for braket notation
\newcommand{\Bk}{\Braket} % shorthand for braket notation
\renewcommand{\v}{\bm} % bold vectors
\renewcommand{\d}{\partial} % partial derivative
\renewcommand{\dd}{\text{d}} % for infinitesimals
\renewcommand{\i}{\mathrm{i}\mkern1mu} % imaginary unit

@@

% symbols with hats
\renewcommand{\b}{\hat b}
\renewcommand{\c}{\hat c}
\newcommand{\s}{\hat s}
\renewcommand{\H}{\hat H}
\renewcommand{\S}{\hat S}
\newcommand{\R}{\hat R}
\newcommand{\T}{\hat T}
\renewcommand{\O}{\hat{\mathcal{O}}}
\renewcommand{\P}{\hat{\mathcal{P}}}

% more shorthands
\newcommand{\up}{\uparrow}
\newcommand{\dn}{\downarrow}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\I}{\mathcal{I}}

\newcommand{\xx}{\x\x}
\newcommand{\xxi}{\x\x_\i}
\newcommand{\X}{\text{X}}
\newcommand{\XX}{\X\X}
\newcommand{\XXI}{\X\X_\i}
\newcommand{\spin}{\text{spin}}
\newcommand{\eff}{\text{eff}}
\newcommand{\crit}{\text{crit}}
\newcommand{\MF}{\text{MF}}
\newcommand{\su}{\mathfrak{su}}
\renewcommand{\ss}{\bar{\v s}\cdot\bar{\v s}}
\DeclareMathOperator{\diag}{diag}

@@

@@

\begin{abstract}
We study multilevel fermions in an optical lattice, with dynamics described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a multilevel spin model with all-to-all SU($n$)-symmetric couplings.
Additionally, Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with the Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU($2$) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU($2$) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless, bentsen2019integrable}.

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, beverland2016realizing, cazalilla2014ultracold, taie2012su, hofrichter2016direct, cappellini2014direct, scazza2014observation, zhang2014spectroscopic, goban2018emergence, perlin2019effective} with exquisite degrees of microscopic control.
As a simple example, this symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes the experimental observations of bosonization and faster thermalization in multi-component Fermi gas with $SU(n)$ symmetry \cite{song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, two-orbital SU($n$) magnetism \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic}, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.
In the spirit of quantum simulation, further investigations in controlled settings will play an important role in understanding the consequence of SU($n$) symmetries for fundamental questions in physics, as well as their practical use in technological applications.
For example, SU($2$)-symmetric spin interactions can be harnessed to develop quantum sensors that surpass classical limits on measurement precision \cite{he2019engineering, perlin2020spin}.
The prospect of similarly exploiting more general SU($n$) symmetries to achieve a technological advantage is still an unexplored avenue of research with untapped potential.

In this work, we consider an experimentally relevant and theoretically tractable regime of the SU($n$) Hubbard model, highlighting differences and similarities with the more familiar case of SU($2$).
Working at ultracold temperatures and unit spatial filling (one atom per lattice site), we begin by mapping the SU($n$) Hubbard model onto a multilevel spin model with all-to-all SU($n$)-symmetric interactions in Section \ref{sec:spin_model}.
We consider the effect of spin-orbit coupling (SOC) in Section \ref{sec:SOC}, finding in particular that the weak-SOC limit generally gives rise to a (synthetic) inhomogeneous magnetic field, extending previously known results to $n>2$ \cite{wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
In Section \ref{sec:controls} we examine the use of control fields to address nuclear spins (and generate SOC), finding a simple three-laser driving scheme that allows for the preparation of interesting states with nontrivial intra-spin correlations when $n>2$.
Finally, we combine these ingredients to examine mean-field dynamical behaviors of the SU($n$) spin model in Section \ref{sec:mean_field}, finding that:
\begin{enumerate*}
\item long-time-averaged observables obey simple scaling relations with the spin dimension $n$, exhibiting (for spin-polarized initial states) dynamical ferromagnetic and dynamical paramegnetic phases, as previously seen for the case of $n=2$ \cite{smale2019observation, lewis-swan2021cavityqed}, and
\item for $n>2$ the long-time dynamics can be highly sensitive to the intra-spin coherences of the initial state.
\end{enumerate*}
We conclude and discuss future directions in Section \ref{sec:conclusions}.

@@

Here we review the realization of a collective SU($n$) spin model in a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.}, \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_k e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

If the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, then the mode-changing collisions in $\H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work)\footnote{Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $\c_{\pi-p,\mu}^\dag \c_{\pi-q,\mu} \c_{p\nu}^\dag \c_{q\nu}$, which conserve both momentum and energy.
We defer a careful treatment of these terms to future work, noting only that they vanish on the manifold of permutationally symmetric spin states with one atom per lattice site.}.
The terms with $k=\ell$ and $p=q$ are $\frac{u}{2N} \sum \c_{\ell\mu}^\dag \c_{\ell\mu} \c_{q\nu}^\dag \c_{q\nu} = \frac12 N u$, which is an overall energy shift that we can freely neglect.
Defining the spin operators $\s_{\mu\nu q}\equiv \c_{q\mu}^\dag \c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians are
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \s_{\mu\mu q},
  \label{H_kin_start} \\
  \H_{\t{int}}
  &= -\f{u}{2N} \sum_{p,q,\mu,\nu} \s_{\mu\nu p} \s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case we can simply treat our system as $N$ distinguishable $n$-level quantum spins at ``lattice sites'' $p,q$.
Note that the ``kinetic'' terms of this spin model ($\H_{\t{kin}}$) are proportional to the identity operator, contributing an overall shift in energy that we can neglect and this point.
Nevertheless, these kinetic terms will become important in the presence of an external drive, which we discuss in Section \ref{sec:SOC}.

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU($2$) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \v\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v{\hat{A}},\v{\hat{B}}$, we define the inner product
\begin{align}
  \v{\hat{A}} \cdot \v{\hat{B}}
  \equiv \sum_{\mu,\nu} A_{\mu\nu}^\dag B_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \v\s_p\cdot\v\s_q
  = -\f{u}{2N}\v\S\cdot\v\S,
  &&
  \v\S \equiv \sum_q \v\s_q.
  \label{eq:H_int}
\end{align}
Here $\v\S$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU($2$) \cite{he2019engineering}, with $\frac12\v\S\cdot\v\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality after modding out identity terms).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U$.
(b,c) When the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, the frozen-mode approximation enables the interaction Hamiltonian to be written as a spin model consisting of exchange terms $\v\s_p\cdot\v\s_q$, which swap the states two spins pinned to momentum modes $p,q$.
(d) Interactions open an energy gap $u=U\times N/L$ between the manifold of permutationally symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v\s_p\cdot\v\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v\s_p\cdot\v\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v\s_p\cdot\v\s_q$ is analogous to the enforcement of SU($2$) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\v\s_p\cdot\v\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

In the case of SU($2$), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

External fields or additional interactions that respect permutational symmetry can induce nontrivial dynamics within the PS manifold.
Moreover, additional terms that explicitly break permutational symmetry can nevertheless lead to interesting dynamics that can be captured within the PS manifold perturbatively, as long as the coupling to non-PS states is weak compared $u$ (see Appendix \ref{sec:pert_theory}) \cite{bravyi2011schrieffer}.
This perturbative regime is thereby efficiently simulable, as the PS manifold has dimension $\sim N^{n-1}$ (as compared to $n^N$ for the entire spin Hilbert space).
Simulating dynamics within the PS manifold requires calculating matrix elements $\bk{\ell|\O|m}$ of spin operators $\O$ with respect to PS states $\ket\ell,\ket{m}$; we discuss this calculation in Appendix \ref{sec:PS_ops}.
